A comprehensive framework to capture the arcana of neuroimaging analysis

### 2018-10-16 manuscript/manuscript.tex
renamed modifications to name_maps
finished reworking paper to include Environment and new pipeline syntax

\lstset{
    basicstyle=\footnotesize,
    captionpos=b,                    % sets the caption-position to bottom
    language=Python,                 % the language of the code
    showstringspaces=false,          % underline spaces within strings only
    stringstyle=\color{stringcolour},     % string literal style
    commentstyle=\color{commentcolour},    % comment style
    keywordstyle=\color{keywordcolour}\bfseries,       % keyword style
}

@@

\begin{abstract}
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

@@

The separation of analysis design and application in the Arcana
framework follows the conceptual divide between classes and objects in
Object-Oriented (OO) software design. \emph{Study} classes encapsulate
imaging data with the suite of analysis methods that can be performed on
them, and, in the basic case, are specialised for a specific imaging
contrast or modality. Study objects apply the analysis suite defined in
the Study class to a specific dataset.

The set of acquired data, the derivatives that can be derived from it,
and methods that construct pipelines to derive the derivatives, are
linked together by the \emph{data specification} \emph{table} of the
Study class (Figure \ref{fig:example_study}). Likewise, free parameters used in pipeline
construction are defined in the class' \emph{parameter specification}
\emph{table}. Class inheritance can be used to specialise analysis
suites by overriding pipeline constructor methods or entries in the
specification tables. Analysis suites for multi-modal data can be
implemented by combining multiple Study classes in \emph{MultiStudy}
classes\emph{.}

@@

Analysis methods are applied to a specific dataset by instantiating an
appropriate Study class and requesting a derivative listed in the class'
data specification table. At instantiation, a Study object is passed
references to \emph{Repository} and \emph{Processor} objects, which
define how and where data is stored and processed. When a derivative is
requested, a Study object queries the repository for intermediate
derivatives that can be reused before constructing a workflow to produce
the requested derivative. Workflows are executed in the computing
environment specified by the Processor object and their products are
stored back in the repository for reuse by subsequent analyses (Figure
\ref{fig:simple_uml}). Input data to the study are also stored in the repository, and are
matched against entries in the class' data specification table using
criteria defined in \emph{Selector} objects passed to the Study object
at instantiation.

@@

\subsubsection*{Data and parameter specification
tables}
\label{data-and-parameter-specification-tables}

@@

Data specified in the data specification table can be of either a
\emph{fileset} or \emph{field} type. Filesets represent single files,
header and data file pairs, or sets of related files typically contained
within single directories (e.g. a multi-volume DICOM dataset). Fields
represent a single integer, floating point value or character string
type. Each Fileset references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

@@

\begin{figure}
\lstinputlisting[language=Python, firstline=2, lastline=58]{listings/example_study.py}
\caption{Example data and parameter specifications. The data
specification specifies two ``acquired'' file sets, `one' and `ten' and
ten derived file sets that can be derived from them, at least
indirectly. Each derived data spec, specifies the name of the pipeline
constructor that creates the pipeline that derives them. Parameter
specifications specify a name and default value for free parameters of
the Study class}
\label{fig:data_spec}
\end{figure}

Both acquired and derived data specifications have a \emph{frequency}
attribute which specifies where the data sits in the hierarchy of the
dataset and can take the values \emph{per\_session},
\emph{per\_subject}, \emph{per\_visit} or \emph{per\_study}. In
addition, derived specifications are passed the name of a method in the
class that constructs the pipeline to derive them. Therefore, while a
pipeline can have multiple outputs, each derivative is derived by only
one pipeline.

@@

\subsubsection*{Workflow design}
\label{workflow-design}

Analysis workflows are implemented in Arcana as a series of pipelines
that each perform a unit of the analysis (e.g. registration, brain
extraction). Each pipeline is in turn constructed from a (typically
small) graph of Nipype nodes, with each node wrapping a stand alone tool
(e.g. FSL's FLIRT) or analysis package function (e.g. SPM's \emph{coreg}
tool).

Pipelines are included in the analysis suite of a Study class by
defining a method in the class that constructs and returns a
\emph{Pipeline} object, and referencing the method by name in a FilesetSpec
or FieldSpec in the class' data specification table. Pipeline objects are thin
wrappers around Nipype workflows, which manage the namespaces of the
workflow nodes, and should be created in constructor methods using the
\emph{pipeline} method of the Study class (Figure \ref{fig:pipeline_constructor}).

\begin{figure}
\lstinputlisting[language=Python, firstline=59]{listings/example_study.py}
\caption{Example pipeline constructor method. Pipelines are
created using the \emph{create\_pipeline} method of the Study class.
Pipeline objects are thin wrappers around Nipype workflows to allow for
namespace management}
\label{fig:pipeline_constructor}
\end{figure}

% Need to introduce and explain new syntax

The syntax for workflow graph construction in Arcana is inspired by the
proposed format for Nipype v2.0
(\href{http://github.com/nipy/nipype/issues/2539}{http://github.com/nipy/nipype/issues/2539}),
and will be updated as Nipype v2.0 is developed and upstream
functionality becomes available. Workflow nodes are created from Nipype
interfaces using the \emph{add(name, interface)} method of the
Pipeline should be used. Once created, the inputs of the node can be set
explicitly via the node's \emph{inputs} attribute, connected to other
nodes with the \emph{connect(out\_node, out\_node\_output,
in\_node, in\_node\_input)} method, or connected to inputs of the
pipeline. Pipeline inputs and outputs are connected to nodes using the
\emph{add\_input(spec\_name, node, node\_input)} and
\emph{add\_output(spec\_name, node, node\_output)} methods,
respectively.

At initialisation, a Pipeline is given a name, which must be unique amongst the
pipelines that can be constructed by the Study. A description of the pipeline and
the list of scientific references

Study class Inputs and outputs refer to entries in the data specification table and
specify the file formats (fileset) or data type (field) that the input
is expected in, or the output will be returned in. If this
format/data-type differs from that specified in the data specification
table then an implicit conversion will be performed when the pipeline is
executed, thereby making data conversion transparent to the researcher.
Every pipeline constructor should take only wildcard keyword arguments
(i.e. **kwargs), which should be passed directly through to the
new\_pipeline method to enable modification of the pipeline namespace by
overriding methods in subclasses.

% End reworked syntax explanation

Any external software packages required by a node should be referenced
in the \emph{requirements} keyword argument as a list of
\emph{Requirement} objects when the node added to the pipeline. These
requirements can be used by the Processor to configure environment
modules \citep{furlani_modules:_1991} if present, to ensure the required software
packages are available. Similarly, the expected memory requirements and
wall time for the node execution should be provided as keyword args,
\emph{memory}, and \emph{wall\_time}, so they can be inserted into job
scripts submitted to schedulers by HPC Processors.

Iteration over subjects and visits is handled implicitly by Arcana and
depends on the frequency of the pipeline's inputs and outputs. To create
a summary derivative (i.e. frequency != \emph{per\_session}) from more
frequent data a join node is used. Join nodes that join over subjects or
visits are created with \emph{add}\_\emph{join\_subjects(name,
interface, joinfield)} and \emph{add\_join\_visits(name,
interface, joinfield)} respectively. As in the underlying Nipype, the
\emph{joinfield} argument specifies which inputs of the receiving node
to join into lists. If direct access to the subject or visit IDs is
required by a node they can be connected to the implicit iterator nodes
with \emph{connect\_subject\_id(node, name)} and
\emph{connect\_visit\_id} respectively. Note that expansion
over iterator sources is handled implicitly by Nipype when nodes with
different iterators are connected to a downstream node.

Study parameters can be accessed during pipeline construction with the
\emph{parameter(name)} method. If conditional logic is included in
the workflow construction that alters the pipeline inputs, outputs or
parameters then it should be controlled by a switch instead of a
parameter. The analysis branch designated by a switch value should be
tested with \emph{branch(name, ref\_value)} method.

@@

Pipeline constructors can be overridden in subclasses like any Python
method. Typically the overriding method will call the superclass method
to construct a base Pipeline, apply modifications and return the
modified Pipeline. In this scenario, references to the data
specification in the superclass method can be mapped onto different
entries in the subclass by passing a dictionary containing the maps to
the \emph{name\_map} keyword argument of pipeline constructor in the
superclass.

Instead of setting the class attributes directly, the data and parameter
specifications are set by the metaclas\emph{s}, in order to combine with
specifications from base classes. The combined data and parameter
specifications are constructed by visiting the class' bases in reverse
method resolution order (MRO) and adding specifications from their
\emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

@@

To apply the analysis blueprint specified in a Study class to a specific
dataset, an instance of the Study class is created with details of where
the data is stored (the \emph{Repository}) and the computing environment
available to process it (the \emph{Processor}). A Study object controls
the construction of analysis workflows, the execution of the workflows,
and the flow of data to and from the repository (Figure \ref{fig:full_uml}).

@@

Each Study object is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. by different
members of a lab). Parameters are set on initialisation of the Study
object along with the range of subject and visit IDs to be included in
the analysis (if they are not provided then all IDs are included). The
remaining arguments passed to the Study object initialisation are the
Repository and Processor modules and a list of \emph{Selector} objects
to select input datasets from the repository (Figure \ref{fig:study_application}).

\begin{figure}
\lstinputlisting[language=Python]{listings/example_application.py}
\caption{Example application of Study class to a dataset stored in an XNAT
repository. Once the Study object has been initialised potential
derivatives of the Study can be requested, and will be generated and
stored in the repository if already present.}
\label{fig:study_application}
\end{figure}

@@

Each study can only have one repository in which derivatives are stored.
However, a study can draw data from multiple auxiliary repositories,
which are specified in the inputs passed to the study. When using
multiple input repositories, subject and visit IDs will often need to be
mapped from their values in the auxiliary repositories to the ``ID
space'' of the study. This mapping is specified using \emph{IdMapper}
objects passed to \emph{subject\_id\_map} and \emph{visit\_id\_map}
keyword arguments during initialisation of a repository.

New repository modules for additional repository types can be
implemented by extending the Repository abstract base class and
implementing a small set of methods (Table \ref{tbl:abstract-repo-methods}).

\begin{table}
\caption{Abstract methods in the base Repository class that
need to be implemented by platform-specific sub-classes.}
\label{tbl:abstract-repo-methods} 
\begin{tabular}{ll}
\hline\noalign{\smallskip}
\textbf{Method name} & \textbf{Function} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\emph{find\_data()} & Queries repository to find all existing filesets and
fields \\
\emph{get\_fileset(Fileset)} & Caches file-set (if necessary) and returns the file path. \\
\emph{get\_field(Field)} & Retrieves and returns the value of the field from
the repository \\
\emph{put\_fileset(Fileset)} & Inserts file-set into the repository and updates cache. \\
\emph{put\_field(Field)} & Inserts the value of the Field into the repository \\
\emph{connect()} & Opens a connection to the repository (optional). \\
\emph{disconnect()} & Closes connection to repository (optional)\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

Selection criteria are specified by list of \emph{FilesetSelector} and
\emph{FieldSelector} objects provided to the \emph{inputs} argument of
the Study initialisation method. Matching is typically performed on file
names (dataset labels for XNAT repositories), and field names. If the
names are inconsistent across the study then regular expressions can be
used instead of exact matches by passing \emph{is\_regex=True} when
initialising the Selector. Additional criteria can be used to
distinguish cases where multiple filesets in the session match the
pattern provided, such as DICOM header values, or when using an XNAT
repository, the order and ID of the dataset.

@@

If the data to select has been derived from an alternative study, then
the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the selector. There are
no restrictions on selecting any data, derived or otherwise, to match
acquired or derived specifications. For example, it is possible to
deliberately skip analysis steps by selecting an output of an early step
in a previous run as a match for a later derived spec although this is
not recommended as standard practice.

Specific files and fields that are not stored within a repository can be
passed as inputs in \emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
table and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
spec). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the ``default
input'' for a data spec using the \emph{default\_input} keyword
argument. For the sake of portability, default inputs should be
restricted to data in publically accessible repositories or included in
standard software packages (e.g. FSL).

@@

Processor modules control how and where pipelines generated by a Study
are executed. There currently four Processor modules implemented in
Arcana: \emph{LinearProcessor}, \emph{MultiProcProcessor},
\emph{SlurmProcessor} and \emph{SgeProcessor}, which wrap the
correspondingly named Nipype execution plugins. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
trivial.

A processor is used internally by a Study instance to execute pipelines
to derive derivatives requested from the data specification by the
\emph{Study.data(name{[}, name,...{]})} method (Figure \ref{fig:simple_uml}). The first
step in this procedure is to query the repository tree for all data and
provenance associated with the study. Sessions for which the requested
outputs of the pipeline are already present in the repository are
excluded from the list to process. For the remaining sessions to
process, inputs of the pipeline that are derivatives themselves are
added to the stack of requested derivatives. This procedure is repeated
recursively until there are no sessions to process or all inputs to the
pipeline are study inputs at a given depth.

When a pipeline is processed it is connected to \emph{Source} and
\emph{Sink} nodes, which get and put the pipeline inputs and outputs
from and to a repository, respectively. Separate source and sink nodes
are used for each data frequency (i.e. per-session, per-subject,
per-visit, per-study). If implicit file format conversion is required
(i.e. the input or output format differs from the data specification)
then additional format converter nodes are inserted after the source
nodes or before the sink nodes. Iterator nodes that iterate over the
required subjects and visits are connected to the sources, and report
nodes that join over subjects and visits are connected to the sink
nodes. Report nodes of upstream pipelines are connected to downstream
iterator nodes to create a single workflow, which is then executed using
the Nipype execution plugin.

Software versions are managed by the processor if environment modules
\citep{furlani_modules:_1991} are installed on the system. These are loaded and
unloaded before the interface of a workflow node is run based on the
requirements specified for the node during construction of the pipeline
(see \emph{Pipeline constructors}. Maps from requirement names to
module names in the processing environment can be passed to the
Processor at initialisation. For plugins that submit jobs to HPC
schedulers (e.g. SLURM, SGE), node memory and wall-time requirements are
inserted into jobscripts where required.

@@

@@

The Arcana framework is distributed as a publicly available software
package via GitHub
(\href{http://github.com/MonashBI/arcana}{{http://github.com/MonashBI/arcana}})
and the Python Package Index (PyPI)
(\href{https://pypi.org/project/arcana/}{{http://pypi.org/project/arcana/}}).
Study classes for T1, T2* and diffusion weighted MRI data have been
implemented as part of the \emph{Biomedical imAgiNg ANAlysis (Banana)} package \\
(\href{https://github.com/MonashBI/banana}{{http://github.com/MonashBI/banana}} and
\href{http://pypi.org./project/banana}{{http://pypi.org./project/banana}}).
All three classes, \emph{T1wStudy}, \emph{T2starwStudy} and
\emph{DmriStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MRIStudy}.
The T1Study and T2starStudy classes are aggregated into the
\emph{T2starwT1wStudy} MultiStudy class to combine analysis of T2* and
T1-weighted contrasts.

@@

The \emph{T2starwT1wStudy} class implements an algorithm to generate
\emph{composite vein images} from the combination of QSM and SWI
contrasts derived from the T2*-weighted acquisition with an vein atlas
registered to the T1-weighted acquisition \citep{ward_combining_2018} (Figure \ref{fig:t2star_t1_study}).

>>>>

\lstset{
    captionpos=b,                    % sets the caption-position to bottom
    language=Python,                 % the language of the code
    showstringspaces=false,          % underline spaces within strings only
    stringstyle=\color{stringcolour},     % string literal style
    commentstyle=\color{commentcolour},    % comment style
    keywordstyle=\color{keywordcolour}\bfseries,       % keyword style
}

@@

\begin{abstract}

Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.

\keywords{Neuroimaging \and Workflows \and Repository \and Large-scale \and Reproducibility}

% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}

\end{abstract}

@@

% Provenanced mentioned here

@@

The separation of analysis design and application in the Arcana
framework follows the conceptual divide between classes and objects in
Object-Oriented (OO) software design. \emph{Study} classes encapsulate
types of data, such as scans of a specific imaging
contrast or modality, with the suite of analysis methods that can be performed on
them. Study objects apply the analysis suite defined in
the Study class to a specific dataset.

The set of acquired data, the derivatives that can be derived from it,
and methods that construct pipelines to derive the derivatives, are
linked together by the \emph{data specification} \emph{table} of the
Study class (Figure \ref{fig:example_study}). Likewise, free parameters used in pipeline
construction are defined in the class' \emph{parameter specification}
\emph{table}. Class inheritance can be used to specialise analysis
suites by overriding entries in the specification tables or pipeline
constructor methods. Analysis suites for multi-modal data can be
implemented by combining Study classes within \emph{MultiStudy}
classes\emph{.}

@@

Analysis methods defined by a Study class are applied to a
specific dataset by instantiating an object of the class and
requesting a derivative listed in the class' data specification table.
At initialisation, a Study object is passed references to a \emph{Repository},
a \emph{Processor}, and an \emph{Environment}, which
define where and how data is stored and processed. When a derivative is
requested, a Study object queries the Repository for intermediate
derivatives that can be reused before constructing a workflow to produce
the requested derivative. The manner in which the workflows are executed
(i.e. single/multi-process or via SLURM scheduler) is specified by the
Processor and software modules required by the analysis are
loaded by the Environment. Selected workflow products are
stored back in the repository for reuse by subsequent analyses (Figure
\ref{fig:simple_uml}). Input data to a study are selected from repositories
 using criteria defined in \emph{Selector} objects passed to the Study
 object at initialisation and matched against entries in the class' data
specification table.

@@

\subsubsection*{Data and parameter specification
tables}
\label{sec:data-and-parameter-specification-tables}

@@

Data specified in the data specification table can be of either a
\emph{fileset} or \emph{field} type. Filesets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point or character string
values. By default, a Field represents a single value but if the \emph{array}
flag is set a Field represents a list of values.
Each Fileset references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

@@

\begin{figure}
\lstinputlisting[basicstyle=\footnotesize,firstline=2, lastline=59]{listings/example_study.py}
\caption{Example data and parameter specifications. The data
specification specifies two ``acquired'' file sets, `one' and `ten' and
ten derived file sets that can be derived from them, at least
indirectly. Each derived data spec, specifies the name of the pipeline
constructor that creates the pipeline that derives them. Parameter
specifications specify a name and default value for free parameters of
the Study class}
\label{fig:data_spec}
\end{figure}

All data specifications have a \emph{frequency} attribute which specifies
where the data sits in the hierarchy of the
dataset and can take the values 'per\_session',
'per\_subject, 'per\_visit' or 'per\_study'. In
addition, derived specifications are passed the name of a method in the
class that constructs the pipeline to derive them. Therefore, while a
pipeline can have multiple outputs, each derivative is derived by only
one pipeline.

@@

\subsubsection*{Workflow design: pipeline constructor methods}
\label{sec:pipeline-constructors}

Workflows are implemented in Arcana as a series of pipelines, which
each perform a unit of the analysis (e.g. registration, brain
extraction, quantitative susceptibility mapping). Pipelines are represented by \emph{Pipeline} objects,
which are thin wrappers around Nipype workflows to handle input/output
connections and namespace management. Each pipeline consists of a
(typically small) graph of Nipype nodes, with each node wrapping a stand alone tool
(e.g. FSL's FLIRT) or analysis package function (e.g. SPM's \emph{coreg}
tool). Pipelines source their inputs from, and sink their outputs to, entries
in the data specification table, thereby linking the table together.

Pipelines are constructed dynamically by ``pipeline constructor" methods
of the study. Pipeline constructor methods are referenced by name in the \emph{pipeline\_name}
argument of the FilesetSpec and FieldSpec objects of data specifications
they derive. Pipeline constructor methods should only receive wildcard keyword
arguments (e.g. \lstinline{my_pipeline(self, **name_maps)}), and these arguments
should be passed as a dictionary directly to the \emph{name\_maps} argument
of the pipeline initialisation to enable inputs and outputs of the
pipeline to be rerouted to alternative data specifications in modified constructor
methods, typically in subclasses and multi-studies.

\begin{figure}
\lstinputlisting[firstline=61, basicstyle=\footnotesize]{listings/example_study.py}
\caption{Example pipeline constructor method. Pipelines are
created using the \emph{pipeline} method of the Study class.
Pipeline objects are thin wrappers around Nipype Workflow objects to
in order manage the namespaces of the workflow's inputs, outputs and nodes. Every
pipeline constructor method should allow wildcard keyword arguments,
which are passed to the \emph{name\_maps} argument of the pipeline
initialisation. This allows pipeline constructors in sub and multi classes to map the
inputs and outputs of the pipeline onto different data specifications.}
\label{fig:pipeline_constructor}
\end{figure}

The syntax for pipeline construction is inspired by a proposed form for Nipype v2.0
(\href{http://github.com/nipy/nipype/issues/2539}{http://github.com/nipy/nipype/issues/2539}).
Within a pipeline constructor, the Pipeline object is initialised by the \emph{pipeline} method
of the Study (Figure \ref{fig:pipeline_constructor}). At initialisation, each pipeline is given a name,
which must be unique amongst the pipelines constructed by the Study. The methods implemented
by the pipeline can be characterised by providing a the list of scientific references
and a text description to the \emph{references} and \emph{desc} keyword arguments, respectively.

The \emph{add(name, interface)} method is used to add a node to a pipeline, which takes
a unique name for the node (within the pipeline) and a Nipype Interface object, and returns
a reference to the newly added node. For clarity, is recommended to put all static inputs
(i.e. parameters) of the interface as keyword arguments of the interface constructor
(Figure \ref{fig:pipeline_constructor}). However, if an input conditionally depends on a parameter of the
study it can be set via the \emph{inputs} attribute of the node (e.g. \lstinline{my_node.inputs.my_param = 1.0}).

Interface traits are connected to inputs and outputs of the pipeline by
providing \emph{inputs} and \emph{outputs} keyword arguments to the \emph{add} method. Both
arguments take a dictionary, where the keys of the dictionary are the names of traits of the interface. The
values of the dictionary are 2-tuples each consisting of the name of a data specification
and the format the pipeline expects the data in (i.e. a FileFormat for Fileset specifications or datatype
for Field specifications). If the expected format does not match that of the corresponding data specification
table or study input then an conversion nodes are implicitly connected to the pipeline.
Inputs and outputs that are conditional on parameters of the study can be set outside of
the \emph{add} method using the \emph{connect\_input} or
\emph{connect\_output} methods, respectively.

Internal connections between nodes within a pipeline are specified by a dictionary
passed to the \emph{connect} keyword argument of the receiving node. The
dictionary keys are the names of the receiving traits and the values are 2-tuples
consisting of the sending node and the name of the sending trait. Alternatively,
conditional connections can be specified via the \emph{connect} method of the
pipeline, which simply calls the method of the same name in the underlying
Nipype workflow.

Any external software packages required by a node should be referenced
in the \emph{requirements} keyword argument as a list of
\emph{Requirement} objects when the node added to the pipeline. These
requirements can be used by the Processor to configure environment
modules \citep{furlani_modules:_1991}, to ensure the required software
packages are available. Similarly, the expected memory requirements and
wall time for the node execution should be provided to the keyword args
\emph{memory}, and \emph{wall\_time} so they can be inserted into HPC
scheduler job scripts if required.

Iteration over subjects and visits is handled implicitly by Arcana and
depends on the frequency of the pipeline's inputs and outputs. To create
a summary derivative (i.e. frequency != 'per\_session') from more frequent
data, \emph{Study.SUBJECT\_ID} or \emph{Study.VISIT\_ID} should be
passed to the \emph{joinsource} keyword to join over subjects or visits, respectively.
In this case, a JoinNode will be created instead of standard Node,
which should be passed the additional keyword argument \emph{joinfield} to
specify the list of input traits to convert into lists to receive the joined input.
Similarly, if the name of an input trait is provided (or list thereof) to the the \emph{iterfield}
keyword argument, then a MapNode will be created and the interface will
be applied to all items of the list connected to that input \citep{gorgolewski_nipype:_2011}.
The subject and visit IDs are directly accessible as input fields of the pipeline
named \emph{Study.SUBJECT\_ID} and \emph{Study.VISIT\_ID}, respectively.

Study parameters can be accessed during pipeline construction with the
\emph{parameter(name)} method. If conditional logic is included in
the workflow construction that alters the pipeline inputs, outputs or
parameters then it should be controlled by a switch instead of a
parameter. The analysis branch designated by a switch value should be
tested with \emph{branch(name)} in the case of boolean switches and
\emph{branch(name, ref\_value)} in the case of string switches.

@@

Instead of being set directly, the data and parameter specifications are set
by the metaclass of the Study (i.e. StudyMetaClass),
in order to combine them with corresponding specifications in the class' bases. The combined
data and parameter specifications are constructed by visiting the class'
bases in reverse method resolution order (MRO) and adding specifications
from their \emph{add\_data\_specs}, \emph{add\_param\_specs} attributes,
overriding previously added specifications with matching names. Note
that in this scheme, specifications can only be appended or overridden
but not removed by Study subclasses so as not to break workflows
inherited from base classes.

Pipeline constructor methods can be overridden in subclasses like any Python
method. Often the overriding method will call the superclass method
to construct a pipeline, apply modifications and return the
modified pipeline. In this scenario, references to the data
specification in the superclass method can be mapped onto different
entries in the table of the subclass by providing the \emph{input\_map} and
\emph{output\_map} keyword arguments when calling the superclass method.
Additionally the name of the pipeline can be altered by providing the \emph{name}
argument, which is useful when creating multiple  constructor methods from the
one base method. In these scenarios it is important to
also pass the the wildcard keyword arguments of the overriding method to the
\emph{name\_maps} keyword argument of the superclass method, to allow the
overriding method to be overridden in turn.

@@

To apply the analysis blueprint specified in a Study class to a specific
dataset, an instance of the Study class is created with details of where
the data is stored (\emph{Repository} module) the computing
resources available to process it (\emph{Processor} module) and the software
installed in the environment it will be processed in (\emph{Environment} module).
A Study object controls the construction and execution of analysis workflows,
and the flow of data to and from the repository (Figure \ref{fig:full_uml}).

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
(if they are not provided then all IDs found in the repository are included).
The remaining arguments passed to the Study object initialisation are the
Repository, Processor and Environment modules to use and a list of
\emph{Selector} objects to select input data from the repository and match
it with the Study's data specification table.
(Figure \ref{fig:study_application}).

\begin{figure}
\lstinputlisting[basicstyle=\footnotesize]{listings/example_application.py}
\caption{Example application of Study class to a dataset stored in an XNAT
repository. Once the Study object has been initialised potential
derivatives of the Study can be requested, and will be generated and
stored in the repository if already present.}
\label{fig:study_application}
\end{figure}

@@

Each study can only have one repository in which derivatives are stored.
However, a study can draw data from multiple auxiliary repositories,
which are specified in the inputs passed to the study. When using
multiple input repositories, subject and visit IDs will often need to be
mapped from their values in the auxiliary repositories to the ``ID
space'' of the study, which can be done by passing either by a dictionary
or callable object (e.g. function) to the \emph{subject\_id\_map} or
\emph{visit\_id\_map} keyword arguments during initialisation of a repository.

New repository modules for additional repository types can be
implemented by extending the Repository abstract base class and
implementing five abstract methods, \emph{find\_data}, \emph{get\_fileset},
\emph{get\_field}, \emph{put\_fileset} and \emph{put\_field}
(Table \ref{tbl:abstract-repo-methods}).

\begin{table}
\caption{Abstract methods in the base Repository class that
need to be implemented by platform-specific sub-classes.}
\label{tbl:abstract-repo-methods} 
\begin{tabular}{ll}
\hline\noalign{\smallskip}
\textbf{Method name} & \textbf{Function} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\emph{find\_data()} & Queries repository to find all existing filesets and
fields \\
\emph{get\_fileset(Fileset)} & Caches fileset (if necessary) and returns the path to file(s). \\
\emph{get\_field(Field)} & Retrieves and returns the value of the field from
the repository \\
\emph{put\_fileset(Fileset)} & Inserts fileset into the repository and updates cache. \\
\emph{put\_field(Field)} & Inserts the value of the Field into the repository \\
\emph{connect()} & Opens a connection to the repository (optional). \\
\emph{disconnect()} & Closes connection to repository (optional)\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

Selection criteria are specified by list of \emph{FilesetSelector} and
\emph{FieldSelector} objects provided to the \emph{inputs} argument of
the Study initialisation method. Matching is typically performed on file
names (dataset labels for XNAT repositories), and field names. If the
names are inconsistent across the study then regular expressions can be
used instead of exact matches by passing \emph{is\_regex=True} when
initialising the Selector. Additional criteria can be used to
distinguish cases where multiple filesets in the session match the
pattern provided, such as DICOM header values, or the order and ID
of the dataset.

@@

If the data to select has been derived from an alternative Arcana Study
instance, then the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the selector. There are
no restrictions on selecting any data, derived or otherwise, to match
acquired or derived specifications. For example, it is possible to
deliberately skip analysis steps by selecting an output of an early step
in a previous run as a match for a later derived spec although this is
not recommended as standard practice.

Specific files and fields that are not stored within a repository can be
passed as inputs in \emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
table and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
spec). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the ``default
input'' for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

@@

Processor modules control how pipelines generated by a Study
are executed. There currently three Processor modules implemented in
Arcana: \emph{LinearProcessor}, \emph{MultiProcProcessor},
\emph{SlurmProcessor}, which wrap the
correspondingly named Nipype execution plugins. The main task performed
by the processor, as separate from the Nipype execution plugin it wraps,
is to determine which pipelines need to be run and link them into a
single workflow. Since this logic is implemented in the Processor
abstract base class, wrapping additional Nipype plugins as required is
trivial.

A Processor is used internally by a Study instance to execute pipelines
to derive derivatives requested from the data specification by the
\emph{data(name{[}, name,...{]})} method (Figure \ref{fig:simple_uml}). The first
step in this procedure is to query the repository tree for all data and
provenance associated with the study. Sessions for which the requested
outputs of the pipeline are already present in the repository, and the stored
provenance matches the current parameters of the study,
excluded from the list to process. For the remaining sessions to
process, inputs of the pipeline that are derivatives themselves are
added to the stack of requested derivatives. This procedure is repeated
recursively until there are no sessions to process or all inputs to the
pipeline are study inputs at a given depth.

When a pipeline is processed it is connected to source and
sink nodes, which get and put the pipeline inputs and outputs
from and to a repository, respectively. Separate source and sink nodes
are used for each data frequency (i.e. per-session, per-subject,
per-visit, per-study). If implicit file format conversion is required
(i.e. the input or output format differs from the data specification)
then additional format converter nodes are inserted after the source
nodes or before the sink nodes. Iterator nodes that iterate over the
required subjects and visits are connected to the sources, and ``deiterator''
nodes that join over subjects and visits are connected to the sink
nodes. Final nodes of upstream pipelines are connected to downstream
iterator nodes in order to create a single workflow, which is then executed using
the Nipype execution plugin.

@@

\subsubsection*{Environment modules}

The software packages installed on the system are detected and managed
by the Environment module. There are currently two types of available Environment modules:
\emph{StaticEnvironment} and \emph{ModulesEnvironment}.

The \emph{StaticEnvironment} module doesn't perform any management, and instead
only checks the existing environment for appropriate software versions before running
the analysis.

The \emph{ModulesEnvironment} module can be used if environment
modules \citep{furlani_modules:_1991} are installed on the system (typical in many
HPC environments). In this case, modules are loaded before the interface of a pipeline node
is run based on the requirements specified for the node during construction of the pipeline
(see \emph{Pipeline constructors} and then unloaded afterwards. Mappings from non-standard
module names installed on the system to requirements required by the Study can be
can be passed to the \emph{packages\_map} keyword argument of the ModulesEnvironment
at initialisation. Likewise non-standard versions for packages can be mapped onto requirements
using the \emph{versions\_map} keyword argument.

@@

The Arcana framework is distributed as a publicly available software
package via GitHub (\url{github.com/MonashBI/arcana}) and the
Python Package Index (PyPI) (\url{pypi.org/project/arcana/}).
Study classes for T1, T2* and diffusion weighted MRI data have been
implemented as part of the \emph{Biomedical imAgiNg ANAlysis (Banana)} package
(\url{github.com/MonashBI/banana}; \url{pypi.org./project/banana}).
All three classes, \emph{T1Study}, \emph{T2starStudy} and
\emph{DmriStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MRIStudy}.
The T1Study and T2starStudy classes are aggregated into the
\emph{T2starwT1Study} MultiStudy class to combine analysis of T2* and
T1-weighted contrasts.

@@

The \emph{T2starwT1Study} class implements an algorithm to generate
\emph{composite vein images} from the combination of QSM and SWI
contrasts derived from the T2*-weighted acquisition with an vein atlas
registered to the T1-weighted acquisition \citep{ward_combining_2018} (Figure \ref{fig:t2star_t1_study}).
